### 2014-05-28 Paper1/paper1.tex
Updated for paper ready-ish to go out.

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Intention can be Determined Despite Disturbances}
}
% Insert Author names, affiliations and corresponding author email.
\\
Justin Horowitz$^{1}$, 
James Patton$^{12\ast}$
\\
\bf{1} Bioengineering, University of Illinois at Chicago, Chicago, Illinois, United States
\\
\bf{2} Sensory Motor Performance Program, Rehabilitation Institute of Chicago, Chicago, Illinois, United States
\\
$\ast$ E-mail: pattonj@uic.edu
\end{flushleft}

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
We derive a novel modeling approach for determining movement intent from force-disturbed arm movement, examine its parameter sensitivity, and validate it on human reaching data. We demonstrate that inverting popular arm modeling approaches algebraically allows determination of movement intent from recordings of force disturbances and arm movement. We then extend this to a more general approach, which reveals strengths, limitations, and underlying assumptions. After populating our model with measured lengths and masses as well as impedance values from the literature, we perform sensitivity analysis to determine the accuracy of the model relative to the uncertainty naturally inherent in point-to-point reaching. We find that sensitivity to parameter inaccuracy is on the order of microns and lower than the inaccuracy of human reaching. Further, most of the sensitivity results from uncertainty with regard to the stiffness of the joints. The robustness of this approach indicated that it was suitable for determining reaching intent from human point-to-point reaching movements disturbed by large force pulses. Intended trajectory showed no change from undisturbed reaching for 150 ms after the onset of disturbance. Beyond this point, stiffening of the arm, change in intent, or both usually lead to deviation from undisturbed reaching. Not all disturbed reaches showed evidence of change in intent. Unlike $\lambda$ models, the reaching intent determined by our model is physiologically plausible. This success allows real-time determination of error that can free learning enhancement and treatment methods like error augmentation and challenge point from needing to prescribe explicit task goals. In addition, extraction from recorded human motion enables direct inspection of and hypothesis testing on the contents of human motor intent. While not explicitly examined, this approach is also valid for determination intent is more complex situations like team sports and national security.

@@

\subsection*{Intent Extraction: Deducing the Desired Trajectory}
We first demonstrate the \textit{intent extraction} approach in human motor control, but we later show that the process is applicable to any controlled process with an invertible highest order plant term. The process begins by presuming a model of the controller. Here, we choose the well-known motion control structure of Shadmehr and Mussa-Ivaldi\cite{shadmehr1994adaptive} where the feedforward aspect of the controller perfectly predicts the plant and linearizes the system through cancellation. Additionally, linear feedback rejects position and velocity error. More complex models can be used without any loss of generality. The equation governing the passive planar dynamics (plant) of musculoskeletal structure is of the form:
\begin{equation}
\underbrace{\overbrace{M(q)\ddot{q}}^{\text{Inertia}}+\overbrace{G(q,\dot{q})}^{\text{Coriolis, Centripal}}}_{\text{Plant}}+E=0
\end{equation}
Where $M$ is the mass matrix, $q$ is the joint angles, $G$ contains both Coriolis and centripetal effects, and $E$ is externally-applied torque that the system might interact with. Dots ($\dot{q}, \ddot{q}$) are used to denote time derivatives of the joint angles. The motion behavior changes with the addition of feedforward and/or feedback controllers:
\begin{equation}
\underbrace{\overbrace{M(q)\ddot{q}}^{\text{Inertia}}+\overbrace{G(q,\dot{q})}^{\text{Coriolis, Centripal}}}_{\text{Plant}}+E=\underbrace{\overbrace{\hat{M}(q_d)\ddot{q}_d}^{\text{Inertia}}+\overbrace{\hat{G}(q_d,\dot{q}_d)}^{\text{Coriolis, Centripal}}+\hat{E}}_{\text{Feedforward Controller}}+\underbrace{K_p(q_d-q)+K_d(\dot{q}_d-\dot{q})}_{\text{Impendance, Feedback Controller}}
\end{equation}
Where a hat ( $\hat{ }$ ) over term indicates the nervous system's best estimate of the planar dynamics, also known as the internal model. This portion of the system serves as an inverse-dynamics feedforward controller that cancels out the dynamics of the arm. If the nervous system has sufficient experience and is expecting $E$, it is included as part of the internal model, $\hat{E}$, otherwise $\hat{E}$ is set to zero. $K_p$ and $K_d$ are the lumped impedance and feedback terms that employ a moving state equilibrium to accomplish the desired trajectory, $q_d$. This $q_d$ signifies the unknown desired trajectory that we seek to discover.

For typical dynamic simulations, in order to determine the trajectory the system (i.e., the forward dynamics problem), this second order differential equation requires standard numerical integration to determine the solution to the initial value problem in time. This first involves algebraic manipulation to solve for $\ddot{q}$, and then integration to determine the state trajectory. The intent extraction approach modifies this method by instead solving for $\ddot{q}_d$:
\begin{equation}
\ddot{q}_d=\hat{M}(q_d)^{-1}\left\{M(q)\ddot{q}+G(q,\dot{q})+E-[\hat{G}(q_d,\dot{q}_d)+\hat{E}+K_p(q_d-q)+K_d(\dot{q}_d-\dot{q})]\right\}
\end{equation}  
$\ddot{q}_d$ can then be integrated to determine the intended state trajectory $q_d(t)$. Provided the model of plant and controller are precise, the initial conditions are available and accurate, the mass matrix estimate $\hat{M}$ is invertible, and the constants are accurate and do not vary, and externally-applied force can be accurately measured, then system yields an accurate estimate of the intent.

It is important to distinguish our intended trajectory, $x_d$, from another variable often considered, the equilibrium point of the muscle actuators. Importantly, this has been claimed to be the intended movement. To consider this, we also generalize our approach to any system with a moving intention that governs the control law. This introduces two additional layers of complexity by separating the plant (arm) into its process (dynamics) and actuator components (muscles) and considering the actuator's equilibrium, $x_e$ (sometimes represented as $\lambda$). The feedforward component must use and share this actuator and hence use the actuator's equilibrium. Without loss of generality, we represent the process $P$ and actuator $A$ as linear (or linearizable) operations in some space, $x$:
\begin{equation}
\underbrace{\overbrace{\sum_{n=1}^N P_nx^{(n)}}^\text{Process}+\overbrace{\sum_{m=1}^M A_m(x-x_e)^{(m)}}^\text{Actuator}}_\text{Plant}+E=0
\end{equation}
The internal model of the process ($\hat{P}$ and $\hat{E}$) predicts system actions, external disturbances, and impedance responses in order to determine $x_e$ that $x$ will track a desired path $x_d$.    
\begin{equation}
\overbrace{\sum_{n=1}^N \hat{P}_n x^{(n)}_d+\hat{E}}^\text{Internal Model}+\overbrace{\sum_{m=1}^M A_m(x_d-x_e)^{(m)}}^\text{Actuator}=0
\end{equation}
We solve this for $x_e$ and substituting into eq 4 gives:
\begin{equation}
\overbrace{\sum_{n=1}^N P_nx^{(n)}}^\text{Process}+E=\overbrace{\sum_{n=1}^N \hat{P}_n x^{(n)}_d+\hat{E}}^\text{Internal Model}+\overbrace{\sum_{m=1}^M A_m (x_d-x)^{(m)}}^\text{Feedback (Actuator)}
\end{equation}

@@

\subsection*{Generation of synthetic data using simple dynamic simulation}
We choose one prevailing arm model, that of Burdet et al. \cite{burdet2006stability}, along with the anatomical mass distribution and geometry parameters of Winter \cite{winter2009biomechanics} to create a model for estimating intent. The model of Burdet et al \cite{burdet2006stability} was adapted to reflect measured physical parameters of the subject who provided the reach. Segment lengths were measured directly while gross body mass was self-reported. These values were converted to segment masses, centers of gravity, and moments of inertia using anatomical landmarks and values from Dempster \cite{dempster1955space} and Winter \cite{winter2009biomechanics}. The muscle torque used to calculate stiffness was approximated by looking backward in time a small amount (500 microseconds). Desired trajectory in time was simulated as a minimum jerk, 5th order polynomial of duration 700ms starting and ending with zero velocity and acceleration.
Simulated reaches were 15 cm long, beginning 38 cm out from and 5.7 cm left of the right shoulder then proceeding right.  We used the following types of perturbing forces for each combination of distance and direction: 
Pulse forces applied in one of the the two directions perpendicular to the direction of movement began when the subject had moved either 10\% or 50\% of the distance to the target and lasted for 150ms.
Noise forces began once the subject moved 3 mm, and lasted for the duration of the motion. The forces were drawn from a white noise generator at 1000 Hz with flat power spectral density of 1N, and then passed through a 4th order low-pass Butterworth filter with cutoff $10\pi$ rad/s.

\subsection*{Error Metric}
Our error metric used deviation in position, where unsigned error was calculated as the mean magnitude of deviation from the nominal trajectory across each of 150 1mm wide bins spaced evenly throughout the 15cm reach.  Mean unsigned error (MUE) was also used to summarize the overall error in each movement for sensitivity analysis. Maximum unsigned error, also called maximum perpendicular distance, was used to measure reaching accuracy to avoid sensitivity to duration of inspection.

@@

\subsection*{Human Subjects}
The human data trajectories analyzed here are drawn from four subjects who gave informed consent in accordance with Northwestern University Institutional Review Board. Two male and two female right-handed subjects (ages 24 to 30) performed the reaches with their right arm and were not compensated. Subjects' arm segment lengths were directly measured \textit{in situ} while body mass was self-reported.

\subsection*{Apparatus}
A planar manipulandum (described in Patton and Mussa-Ivaldi \cite{patton2004robot}) was programmed to compensate and minimize any handle friction or mass. The MATLAB XPC-TARGET package \cite{MATLAB:2008} was used to render this force environment at 1000 Hz and data were collected at 1000 Hz.  Visual feedback was performed at 60 Hz using OpenGL. Closed loop data transmission time (position measurement to completed rendering to recognition of rendering by the position measurement system) was less than 8 ms, ensuring a visual delay less than one 60 Hz frame.

\subsection*{Protocol}
Subjects made 730 reaches in total, along a line parallel to their coronal plane and approximately 45 cm from their shoulder. Reaches were either 15 or 30 cm long, starting and ending at one of three points spaced 15 cm apart on the line. To prevent any learning effect, forces were presented intermittently with random frequency, but never less than 5 reaches apart. Forces were chosen pseudorandomly such that each type, direction, and distance combination mentioned above was presented 5 times.

\subsection*{Statistical Analysis}
The MATLAB statistics toolbox package \cite{MATLAB:2008} was used to perform comparisons among subjects and trajectory sources using the nonparametric Kruskal-Wallis test with Tukey-Kramer post-hoc for group differences. Nonparametric testing was necessary due to floor effects from using unsigned measures.

@@

\subsection*{Model Robustness}
Sensitivity analysis fortunately revealed that variance in intended trajectories due to estimated uncertainty in model parameters is lower than the natural variation in undisturbed motions.  Simulated point-to-point reaches disturbed by either filtered white noise forces or a pulse force perpendicular to the direction of the reach (figure 1) were extracted in the presence of 220,000 variations upon the parameters (table 1) according to the methods of Saltelli et al. \cite{saltelli2010variance}. Expected variance due to parameter uncertainty was $0.86  \mu m^2$ for pulse forces and $1.5 \mu m^2$ for filtered white forces. Variance in recorded undisturbed point-to-point reaching under the same time and reach distance conditions was $2.93 \mu m^2$. These variances all fall beneath the instantaneous accuracy of the robot recording the motion (tenths of a millimeter). In other words, while some inversion processes might be highly sensitive to model uncertainty, our process for recovering intent from action is much more precise than its input.

\subsection*{Parameter Sensitivity}
Simulated error due to direct parameter uncertainty (figure 2) reached the order of millimeters and revealed particular sensitivity to misestimation of stiffness and changes in shoulder position from trial to trial. Indirect sensitivity, which reveals how errors from other parameters are affected by changes in a given parameter, was an order of magnitude lower than direct sensitivity and again implicated stiffness estimation inaccuracy as a primary cause of extraction uncertainty. Note that this sensitivity analysis does not provide information on whether models or parameter estimates are accurate, only how sensitive the model would be if they were inaccurate.

\subsection*{Human Intent Trajectories}
Extraction of intent from human point-to-point reaching revealed straight-line movement that persisted for hundreds of milliseconds after the onset of disturbing forces (figure 3). In both pulse timing conditions, intent remained straight line towards the target, supporting the approach.  When disturbances occurred early in reaching, intent showed some signs of correction mid-reach. When disturbances occurred late in reaching, intent continued onto the target (figure 3 lower plots). We observed a systematic change of intent 150 milliseconds after the onset of force disturbance. While the hand returns within 600 milliseconds to undisturbed levels of error, intention is altered for almost 900 milliseconds (figure 4 shaded areas).

As this is the first sound method for looking at intent following disturbance, we have only the undisturbed intention for comparison. If intent can change following disturbance, it can only do so after some period of time due to delay in processing sensory information and producing new motor commands. This represents a way to falsify the method: our determined intent should not diverge from the undisturbed intent of similar motions within this window of delay. As demonstrated in figure 4, the method passes this test until 150 milliseconds after the onset of disturbance.

\section*{Discussion}
We sought to test the suitability and robustness of an algorithm that determines a person's intent during reaching motion, even if there disturbances. Sensitivity analysis on synthetic data revealed that the method was suitable in that errors in response to parameter variations were smaller than trial-to-trial variance observed in human reaching. Significant misestimation could occur if the stiffness was inaccurately modeled. We tested this on human reaching and found that in spite of the hand being distubed by forces, the intended movement remained on-course to the target until 120-150 ms before responding. This represents a new and accurate method for viewing intent and how it changes when faced with a force disturbance.

@@

Not only is intent slow to change, it does not necessarily change following disturbance. While this work examines the time of onset of change, statistical testing cannot be used to detect the absence of change in an individual movement. Still, figure 3 shows reaches that are more compatible with intermittent control than with continuous optimal feedback control. In classic optimal control, the system is always updating the intent in response to errors \cite{todorov2002optimal}. Our results suggest a latent response consistent with sensory feedback latencies \cite{pruszynski2012optimal}. Interestingly, disturbances that come later in the movement (figure 3, bottom panel) appeared to not dramatically affect some subjects' intent, suggesting that people may stubbornly ``stick to their guns'' when they know that the stiffness and damping characteristics will provide enough restoring torque to arrive at the target. Time of response may be dependent on the time of the disturbance in the movement. Alternatively, control might be intermittent \cite{gawthrop2011intermittent} with limited opportunities to change intent. It remains to be seen if and how change in intent might be triggered.

@@

Determination of intended trajectories from force-disturbed movements allows hypothesis testing that was not previously possible. It is important to reiterate that because the human arm's mass is almost always invertible, this approach can take \textbf{any} ``candidate model'' of the human arm and create a filter that estimates the intended trajectory. By using tests such as the comparison with undisturbed straight line reaching employed in this paper, one can evaluate which candidate produces the most plausible intended trajectory, supporting one model over another.  Instead of comparing two generative models to see which better approximates movement, it is now possible to perform element-by-element fitting. Mid-movement replanning is now falsifiable: do intended trajectories from disturbed motions differ from undisturbed motions? They do. Similarly, compositional claims, such as whether movements are comprised of subunits, should now be directly testable. Finally, optimal feedback control models can now gain direct access to an intended trajectory to better understand and fit hypothesized cost functions.

The process model does not need to be explicitly formulated. Dynamic models that strive to understand intent are wide ranging, and can include mathematical models that range in application from national security to athletic performance. The plant (such as the actions of a crowd in a crowded square or the motions of an opposing team) may be too complex to model with simple linear transformations as we present above. However, by substituting some approximation or lookup table of input-outcome pairs, it is possible in a variety of applications to obtain the intent, even when disturbed. It is not necessary to allow a bomber to succeed or an opposing team to be victorious to understand the plan behind their action.

@@

% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=4in]{fig1.eps}
\end{center}
\caption{
{\bf Simulated data illustrating tautology of extraction absent parameter error across pulse and filtered Gaussian noise disturbance types.} Intent is modeled as a minimum jerk, 5th order polynomial. Forces experienced are combined with intent via Burdet et al.'s \cite{burdet2006stability} model to produce the simulated arm trajectory. Extraction to recover intention from arm and force trajectory follows.
}
\label{fig:synthetic}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=4in]{fig2.eps}
\end{center}
\caption{
{\bf Model sensitivity testing (from the variance-based method of Saltelli et al.\cite{saltelli2010variance}) reveals that the model is mostly sensitive to stiffness.} Parameter values are varied in a Sobol-distributed fashion according to Table 1. First order sensitivity, in purple, is the error that would be removed if the parameter was fixed at its nominal value. Total sensitivity, in teal, shows error that would remain if all other parameters were fixed at their nominal values.
}
\label{fig:sensitivity}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=4in]{fig3.eps}
\end{center}
\caption{
{\bf Subjects' disturbed hand trajectories and extraction of intended trajectories from them reveal that both variance and invariance of the motor plan can occur even in response to very large disturbances.} The hand path, in black, deviates from the green baseline, as a force pulse, gray arrows, are applied. This subject's extracted desired trajectories, in red (baseline in faded red), do not significantly differ from baseline during force application, but show some reaction afterwards.
}
\label{fig:anecdotes}
\end{figure}

>>>>

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Determining the Intent of Action Despite Disturbances}
}
% Insert Author names, affiliations and corresponding author email.
\\
Justin Horowitz$^{1}$, 
James Patton$^{1,2\ast}$
\\
\bf{1} Bioengineering, University of Illinois at Chicago, Chicago, Illinois, United States
\\
\bf{2} Sensory Motor Performance Program, Rehabilitation Institute of Chicago, Chicago, Illinois, United States
\\
$\ast$ E-mail: pattonj@uic.edu
\end{flushleft}

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
We derive a novel modeling approach for determining movement intent from a force-disturbed planar arm trajectory, examine its parameter sensitivity, and test it on human reaching data. Popular modeling approaches can be inverted in a general approach, and we show here the strengths, limitations, and underlying assumptions. We found that sensitivity to parameter inaccuracies was on the order of microns and hence much lower than the inaccuracies seen in human targeted reaching. Further, most of the sensitivity resulted from uncertainty with regard to the joint stiffness. We also tested the robustness of this approach on human point-to-point reaching movements disturbed by large force pulses. As expected, the intended trajectory showed no change from undisturbed reaching for 150 ms after the disturbance onset. Beyond this point, intent change, arm stiffening, or both usually lead to deviation from undisturbed reaching. Not all disturbed reaches showed evidence of change in intent. Unlike $\lambda$ models, the reaching intent determined by our model is physiologically plausible. This success allows real-time determination of error that can free learning enhancement and treatment methods like error augmentation and challenge point from needing to prescribe explicit task goals. In addition, extraction from recorded human motion enables direct inspection of and hypothesis testing on the contents of human motor intent. While not explicitly examined, this approach is also valid for determining intent in more complex situations such as human-machine interaction, team sports, or national security.

@@

\subsection*{Intent Extraction: Deducing the Desired Trajectory}
We first demonstrate the \textit{intent extraction} approach in human motor control, but we later show that the process is applicable to any controlled process with an invertible highest order plant term. The process begins by presuming a model of the controller. Here, we choose the well-known motion control structure of Shadmehr and Mussa-Ivaldi\cite{shadmehr1994adaptive} where the feedforward aspect of the controller perfectly predicts the plant and linearizes the system through cancellation. Additionally, linear feedback rejects position and velocity error. This model is chosen to be very simple to help illuminate the approach, but it is limited in how it captures all of the complexity of motor control. Nevertheless later we will discuss how this might be used on more complex models without loss of generality. The equation governing the passive planar dynamics (plant) of musculoskeletal structure is of the form:
\begin{equation}
\underbrace{\overbrace{M(q)\ddot{q}}^{\text{Inertia}}+\overbrace{G(q,\dot{q})}^{\text{Coriolis, Centripal}}}_{\text{Plant}}+E=0
\end{equation}
Where $M$ is the mass matrix, $q$ is the joint angles, $G$ contains both Coriolis and centripetal effects, and $E$ is externally-applied torque that the system might interact with. Dots ($\dot{q}, \ddot{q}$) are used to denote time derivatives of the joint angles. The motion behavior changes with the addition of feedforward and/or feedback controllers:
\begin{equation}
\underbrace{\overbrace{M(q)\ddot{q}}^{\text{Inertia}}+\overbrace{G(q,\dot{q})}^{\text{Coriolis, Centripal}}}_{\text{Plant}}+E=\underbrace{\overbrace{\hat{M}(q_d)\ddot{q}_d}^{\text{Inertia}}+\overbrace{\hat{G}(q_d,\dot{q}_d)}^{\text{Coriolis, Centripal}}+\hat{E}}_{\text{Feedforward Controller}}+\underbrace{K_p(q_d-q)+K_d(\dot{q}_d-\dot{q})}_{\text{Impendance, Feedback Controller}}
\end{equation}
Where a hat ( $\hat{ }$ ) over term indicates the nervous system's best estimate of the planar dynamics, also known as the internal model. This portion of the system serves as an inverse-dynamics feedforward controller that cancels out the dynamics of the arm. If the nervous system has sufficient experience and is expecting $E$, it is included as part of the internal model, $\hat{E}$, otherwise $\hat{E}$ is set to zero. $K_p$ and $K_d$ are the lumped impedance and feedback terms that employ a moving state equilibrium to accomplish the desired trajectory, $q_d$. This $q_d$ signifies the unknown desired trajectory that we seek to discover.

For typical dynamic simulations, in order to determine the trajectory of the system (i.e., the forward dynamics problem), this second-order differential equation requires standard numerical integration to determine the solution to the initial value problem in time. This first involves algebraic manipulation to solve for $\ddot{q}$, and then integration to determine the state trajectory. The intent extraction approach modifies this method by instead solving for $\ddot{q}_d$:
\begin{equation}
\ddot{q}_d=\hat{M}(q_d)^{-1}\left\{M(q)\ddot{q}+G(q,\dot{q})+E-[\hat{G}(q_d,\dot{q}_d)+\hat{E}+K_p(q_d-q)+K_d(\dot{q}_d-\dot{q})]\right\}
\end{equation}  
$\ddot{q}_d$ can then be integrated to determine the intended state trajectory $q_d(t)$. Provided the model of plant and controller are precise, the initial conditions are available and accurate, the mass matrix estimate $\hat{M}$ is invertible, and the constants are accurate and do not vary, and externally-applied force can be accurately measured, then the system yields an accurate estimate of the intent.

It is important to distinguish our intended trajectory, $x_d$, from another variable often considered, the equilibrium point of the muscle actuators. Importantly, this has been claimed to be the intended movement. To consider this, we also generalize our approach to any system with a moving intention that governs the control law. This introduces two additional layers of complexity by separating the plant (arm) into its process (dynamics) and actuator components (muscles) and considering the actuator's equilibrium, $x_e$ (sometimes represented as $\lambda$). The feedforward component must use and share this actuator and hence use the actuator's equilibrium. Without loss of generality, we represent the process $P$ and actuator $A$ as linear (or linearizable) operations in some space, $x$:
\begin{equation} \label{eq:generalPlant}
\underbrace{\overbrace{\sum_{n=1}^N P_nx^{(n)}}^\text{Process}+\overbrace{\sum_{m=1}^M A_m(x-x_e)^{(m)}}^\text{Actuator}}_\text{Plant}+E=0
\end{equation}
The internal model of the process ($\hat{P}$ and $\hat{E}$) predicts system actions, external disturbances, and impedance responses in order to determine $x_e$ that $x$ will track a desired path $x_d$.    
\begin{equation}
\overbrace{\sum_{n=1}^N \hat{P}_n x^{(n)}_d+\hat{E}}^\text{Internal Model}+\overbrace{\sum_{m=1}^M A_m(x_d-x_e)^{(m)}}^\text{Actuator}=0
\end{equation}
We solve this for $x_e$ and substituting into \eqref{eq:generalPlant} gives:
\begin{equation}
\overbrace{\sum_{n=1}^N P_nx^{(n)}}^\text{Process}+E=\overbrace{\sum_{n=1}^N \hat{P}_n x^{(n)}_d+\hat{E}}^\text{Internal Model}+\overbrace{\sum_{m=1}^M A_m (x_d-x)^{(m)}}^\text{Feedback (Actuator)}
\end{equation}

@@

\subsection*{Generation of synthetic data using simple dynamic simulation}
We choose one prevailing arm model, that of Burdet et al. \cite{burdet2006stability}, along with the anatomical mass distribution and geometry parameters of Winter \cite{winter2009biomechanics} to create a model for estimating intent. The model of Burdet et al \cite{burdet2006stability} was adapted to reflect measured physical parameters of the subject who provided the reach. Segment lengths were measured directly while gross body mass was self-reported. These values were converted to segment masses, centers of gravity, and moments of inertia using anatomical landmarks and values from Dempster \cite{dempster1955space} and Winter \cite{winter2009biomechanics}. The muscle torque used to calculate stiffness was approximated by looking backward in time a small amount (500 microseconds). Desired trajectory in time was simulated as a minimum jerk, 5th order polynomial of duration 700 milliseconds starting and ending with zero velocity and acceleration.

Simulated reaches were 15 centimeters long, beginning 38 centimeters out from and 5.7 centimeters left of the right shoulder then proceeding right.  We used the following types of perturbing forces for each combination of distance and direction: 
pulse forces applied in one of the the two directions perpendicular to the direction of movement began when the subject had moved either 10\% or 50\% of the distance to the target and lasted for 150ms.
Noise forces began once the subject moved 3 millimeters, and lasted for the duration of the motion. The forces were drawn from a white noise generator at 1000 Hz with flat power spectral density of 1N, and then passed through a 4th order low-pass Butterworth filter with cutoff $10\pi$ rad/s.

\subsection*{Error Metric}
Our error metric used deviation in position, where unsigned error was calculated as the mean magnitude of deviation from the nominal trajectory across each of 150 bins measuring 1 millimeter in width and spaced evenly throughout the 15centimeters reach.  Mean unsigned error (MUE) was also used to summarize the overall error in each movement for sensitivity analysis. Signed error relative to the direction of disturbance, also called perpendicular deviation, was used to measure reaching accuracy in order to elucidate direction of any corrections.

@@

\subsection*{Human Subjects}
The human data trajectories analyzed here are drawn from four subjects who gave informed consent in accordance with Northwestern University Institutional Review Board, which specifically approved this study and follows the principles expressed in the Declaration of Helsinki. Two male and two female right-handed subjects (ages 24 to 30) performed the reaches with their right arm and were not compensated. Subjects' arm segment lengths were directly measured \textit{in situ} while body mass was self-reported.

\subsection*{Apparatus}
A planar manipulandum (described in Patton and Mussa-Ivaldi \cite{patton2004robot}) was programmed to compensate and minimize any handle friction or mass. The MATLAB XPC-TARGET package \cite{MATLAB:2008} was used to render this force environment at 1000 Hz and data were collected at 1000 Hz.  Visual feedback was performed at 60 Hz using OpenGL. Closed-loop data transmission time (position measurement to completed rendering to recognition of rendering by the position measurement system) was less than 8 milliseconds, ensuring a visual delay less than one 60 Hz frame.

\subsection*{Protocol}
Subjects made 730 reaches in total, along a line parallel to their coronal plane and approximately 45 centimeters from their shoulder. Reaches were either 15 or 30 centimeters long, starting and ending at one of three points spaced 15 centimeters apart on the line. To prevent any learning effect, forces were presented intermittently with random frequency, but never less than 5 reaches apart. Forces were chosen pseudorandomly such that each type, direction, and distance combination mentioned above was presented 5 times.

\subsection*{Statistical Analysis}
The MATLAB statistics toolbox package \cite{MATLAB:2008} was used to perform comparisons among subjects and trajectory sources using the nonparametric Kruskal-Wallis test with Tukey-Kramer post-hoc for group differences. Nonparametric testing was necessary due to differences in the shapes of the force-disturbed and undisturbed hand error distributions.

@@

\subsection*{Model Robustness}
Sensitivity analysis fortunately revealed that variance in intended trajectories due to estimated uncertainty in model parameters is lower than the natural variation in undisturbed motions.  Simulated point-to-point reaches disturbed by either filtered white noise forces or a pulse force perpendicular to the direction of the reach (Figure 1) were extracted in the presence of 220,000 variations upon the parameters (Table 1) according to the methods of Saltelli et al. \cite{saltelli2010variance}. Expected variance due to parameter uncertainty was $0.86  \mu m^2$ for pulse forces and $1.5 \mu m^2$ for filtered white forces. Variance in recorded undisturbed point-to-point reaching under the same time and reach distance conditions was $2.93 \mu m^2$. These variances all fall beneath the instantaneous accuracy of the robot recording the motion (tenths of a millimeter). In other words, while some inversion processes might be highly sensitive to model uncertainty, our process for recovering intent from action is much more precise than its input.

\subsection*{Parameter Sensitivity}
Simulated error due to direct parameter uncertainty (Figure 2) reached the order of millimeters and revealed particular sensitivity to misestimation of stiffness and changes in shoulder position from trial to trial. Indirect sensitivity, which reveals how errors from other parameters are affected by changes in a given parameter, was an order of magnitude lower than direct sensitivity and again implicated stiffness estimation inaccuracy as a primary cause of extraction uncertainty. Note that this sensitivity analysis does not provide information on whether models or parameter estimates are accurate, only how sensitive the model would be if they were inaccurate.

\subsection*{Human Intent Trajectories}
Extraction of intent from human point-to-point reaching revealed straight-line movement that persisted for hundreds of milliseconds after the onset of disturbing forces (Figure 3). In both pulse timing conditions, intent remained straight line towards the target, supporting the approach.  When disturbances occurred early in reaching, intent showed some signs of correction mid-reach. When disturbances occurred late in reaching, intent continued onto the target (Figure 3 lower plots). We observed no significant change of intent during force disturbance relative to undisturbed reaching. Despite statistical insignificance, intention showed some systematic deflection opposite the direction of disturbance beginning around 275 milliseconds after the onset of force disturbance. While the hand returns within 600 milliseconds to undisturbed levels of error, intention appears to over-correct and is altered for almost 900 milliseconds (Figure 4 shaded areas).

As this is the first sound method for looking at intent following disturbance, we have only the undisturbed intention for comparison. If intent can change following disturbance, it can only do so after some period of time due to delay in processing sensory information and producing new motor commands. This represents a way to falsify the method: our determined intent should not diverge from the undisturbed intent of similar motions within this window of delay. As demonstrated in Figure 4, the method passes this test.

\section*{Discussion}
We sought to test the suitability and robustness of an algorithm that determines a person's intent during reaching motion, even if there are disturbances. Sensitivity analysis on synthetic data revealed that the method was suitable in that errors in response to parameter variations were smaller than trial-to-trial variance observed in human reaching. Significant misestimation could occur if the stiffness was inaccurately modeled. We tested this on human reaching and found that in spite of the hand being distubed by forces, the intended movement remained on-course to the target, perhaps changing between 200 and 300 milliseconds. This represents a new and accurate method for viewing intent and how it changes when faced with a force disturbance.

@@

Intent is slow to change if it changes at all following disturbance. While this work examines the time of onset of change, statistical testing cannot be used to detect the absence of change in an individual movement. Still, Figure 3 shows reaches that are more compatible with intermittent control than with continuous optimal feedback control. In classic optimal control, the system is always updating the intent in response to errors \cite{todorov2002optimal}. Our results suggest a latent response consistent with sensory feedback latencies \cite{pruszynski2012optimal}. Interestingly, disturbances that come later in the movement (Figure 3, bottom panel) appeared to not dramatically affect some subjects' intent, suggesting that people may stubbornly ``stick to their guns'' when they know that the stiffness and damping characteristics will provide enough restoring torque to arrive at the target. Time of response may be dependent on the time of the disturbance in the movement. Alternatively, control might be intermittent \cite{gawthrop2011intermittent} with limited opportunities to change intent. It remains to be seen if and how change in intent might be triggered.

@@

Determination of intended trajectories from force-disturbed movements allows hypothesis testing that was not previously possible. It is important to reiterate that because the human arm's mass is almost always invertible, this approach can take \textbf{any} ``candidate model'' of the human arm and create a filter that estimates the intended trajectory. By using tests such as the comparison with undisturbed straight line reaching employed in this paper, one can evaluate which candidate produces the most plausible intended trajectory, supporting one model over another.  Instead of comparing two generative models to see which better approximates movement, it is now possible to perform element-by-element fitting. Mid-movement replanning is now falsifiable: do intended trajectories from disturbed motions differ from undisturbed motions? Not always. Similarly, compositional claims, such as whether movements are comprised of subunits, should now be directly testable. Finally, optimal feedback control models can now gain direct access to an intended trajectory to better understand and fit hypothesized cost functions.

The process model does not need to be explicitly formulated. Dynamic models that strive to understand intent are wide ranging, and can include mathematical models that range in application from national security to athletic performance. The plant (such as the actions of a crowd in a crowded square or the motions of an opposing team) may be too complex to model with simple linear transformations as we present above. However, by substituting some approximation or lookup table of input-outcome pairs, it is possible in a variety of applications to obtain the intent, even when disturbed. In other words, it may not be necessary to allow a bomber to succeed or an opposing team to be victorious to understand the plan behind their action.

@@

% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}
Research supported by NIH NS053606.  We thank Felix Huang, Jagan Jimmy, Christine Massie, Tejas Madhavan, and Mohi Ahmed for their review of preliminary manuscript. We also the community  of the Robotics Lab at the Rehabilitation Institute of Chicago for critical commentary during the genesis of this paper and the work leading up to it.

@@

\begin{figure}[!ht]
\begin{center}
%\includegraphics[width=4in]{fig1.eps}
\end{center}
\caption{
{\bf Simulated data illustrating tautology of extraction absent parameter error across pulse and filtered Gaussian noise disturbance types.} Intent is modeled as a minimum jerk, 5th order polynomial. Forces experienced are combined with intent via Burdet et al.'s \cite{burdet2006stability} model to produce the simulated arm trajectory. Extraction to recover intention from arm and force trajectory follows.
}
\label{fig:synthetic}
\end{figure}

\begin{figure}[!ht]
\begin{center}
%\includegraphics[width=4in]{fig2.eps}
\end{center}
\caption{
{\bf Model sensitivity testing (from the variance-based method of Saltelli et al.\cite{saltelli2010variance}) reveals that the model is mostly sensitive to stiffness.} Parameter values are varied in a Sobol-distributed fashion according to Table 1. First order sensitivity, in purple, is the error that would be removed if the parameter was fixed at its nominal value. Total sensitivity, in teal, shows error that would remain if all other parameters were fixed at their nominal values.
}
\label{fig:sensitivity}
\end{figure}

\begin{figure}[!ht]
\begin{center}
%\includegraphics[width=4in]{fig3.eps}
\end{center}
\caption{
{\bf Subjects' disturbed hand trajectories and extraction of intended trajectories from them reveal that both variance and invariance of the motor plan can occur even in response to very large disturbances.} The hand path, in black, deviates from the green baseline, as a force pulse, gray arrows, are applied. This subject's extracted desired trajectories, in red (baseline in faded red), do not significantly differ from baseline during force application, but show some reaction afterwards.
}
\label{fig:anecdotes}
\end{figure}
